A generalization of the Bradley-Terry model for draws in chess

### 2020-09-15 inst/jebo_chess_rebuttal.txt

An excellent observation.  In the chess world the pre-eminent strength
measure would be the Elo rating in which the winning player takes
points from the loser based on the pre-match difference in Elo rating.
Elo exhibits an emergent phenomenon in which sudden changes of ability
are not reflected instantly but have an exponentially damped response.

My view is that this damping feature is sometimes desirable (old games
are surely less informative than more recent ones) but uncontrolled in
the sense that the time-constant is not explicit or controllable, and
the Elo system uses an arbitrary points system with arbitrary cut-offs
[as opposed to the objective Bradley-Terry system].

But frankly I don't want to open the can of worms that would be
comparison between Elo and Bradley-Terry.  I had a protracted argument
on this issue with a chess enthusiast; let us say the discussion was
intense and robust.  The problem is that Elo was designed specifically
to downweight old data and this downweighting is what users of such
metrics want.  Unfortunately, what my correspondent wanted from BT was
pretty much Elo: a somewhat circular position.

@@

> 
> Precise estimation will hinge on a large number of games, but this
> will be in tension with the assumption of time-invariant
> player-strengths.  The author should explore this tension.
>

I am in total agreement with this observation.  My thinking on this
issue has been strongly affected by my recent work applying
Bradley-Terry to Formula 1 motor racing.  In F1, drivers generally
start their careers as weak competitors, become stronger with
experience, and finally become uncompetitive due to age [although
frankly Lewis Hamilton at 35yo seems to buck this general trend, as
recent events in Tuscany would indicate].  I would suggest that chess
players have a similar career arc, although Formula 1 Teams typically
take a dim view of underperforming drivers---whom they rapidly and
unceremonially drop.  The career arc for these competitors would have
a truncated right hand side, unlike chess players who typically
decline gradually over decades [although I would observe that
Kasparov's current FIDE rating of 2812 is not that much less than his
peak of 2851].

Specifically, the dataset in question stretches from 1981 to 2017, a
very wide range.  Generalized Bradley-Terry, together with reified
entities, is capable of handling strengths varying over a career: one
would introduce a "mid-career monster", or perhaps an "experience
monster".  I would add here that monsters' strength is inherently
positive, so one could deal with waning strength with age by
introducing a "youth monster" who helps younger players.  But such
analysis is, I would suggest, firmly in the category of further work.
I have added a discussion of this issue to the manuscript (RED), in
two parts: firstly when introducing the dataset, and secondly under
Further Work.

> 2) The author neglects to motivate the paper beyond its application
> within the chess world.  The tournament literature's applications lie
> primarily, though definitely not exclusively, within compensation and
> labor.  How could this extension be used within that context?  Where
> might economists see the first-mover advantage in action? Where might
> the extension to allow draws be especially lucrative? Etc.
>

> 3) As the author notes, the work of Moul and Nye (2009) is directly on
> point.  The author's discussion of this paper, however, is overly
> sparse and at times misleading.  Moul and Nye use existing observed
> player strengths (variations of commonly used scores based on the Elo
> rating system that goes back to 1960).  Moul and Nye then use these
> strengths within an ordered probit to estimate parameters that relate
> back to draw likelihood and White's first-mover advantage.  The sharp
> difference between that paper and this is then that the author is
> estimating player relative strengths (conditional upon a pool of
> players) while Moul and Nye use "observed" strengths [footnote: Those
> relative strengths are highly predictive, with t-statistics of around
> 20].  The latter approach is far less parametrically demanding, but
> leans much more on the assumption of normally distributed disturbances
> (basically the differing assumption of functional form).  The latter
> approach also allows individuals' player strength to vary across time
> and is far more flexible to changes in the composition of the pool of
> players.
> 
> Furthermore, Moul and Nye estimate several versions of their ordered
> probit without data on the number of moves (ply-count).  This is
> contrary to the implication of the paper (p. 6, immediately below
> Table 2), which suggests that the ply-count data are essential to that
> paper's findings.  All of that paper's core results (significance of
> draws, first-mover advantage, a distinct draw outcome for round-robin
> international tournament games that involve two Soviets) are
> originally derived without ply-count data.  The ply-count data are
> incorporated to explore a potential mechanism by which players in a
> draw-cartel might benefit.
> 
> 4) The paper concludes that the estimates derived from the 1962
> Interzonals and the 1963 Candidates strongly suggest that the "draw
> monster" won games involving two Soviet players more frequently than
> games that involved at least one non-Soviet.  This is consistent with
> a Soviet cartel, but it could also be the result of Soviet-specific
> styles of play.  Moul and Nye (2009) test the Soviet style-of-play
> hypothesis by considering international tournaments when the format
> changed from all-play-all to knockout and all Soviet tournaments.
> Both comparisons yield results consistent with a Soviet draw-cartel in
> international all-play-all tournaments and rejecting the Soviet
> style-of-play hypothesis.  By only considering tournaments that were
> international and all-play-all, the author forgoes this critical
> source of identification.
> 
> In the cartel example, the author also neglects the advantage that
> presumably accrued to Soviet players when they played non-Soviet
> players in all-play-all international tournaments.  This would suggest
> another "monster" to be estimated, one that benefits the Soviet player
> when playing a non-Soviet opponent.  Without such an addition, Soviet
> strengths will be overstated.
>

### 2020-09-15 inst/jebo_chess_revision2.tex
discussion of Elo rating
* inst/jebo_chess_rebuttal.txt:

\textcolor{red}{One strong assumption made in the
  Kasparov-Karpov-Anand analysis was the constancy of the players'
  strengths over a 36-year period.  Many sports have a career arc
  whereby competitors start their careers weak, become stronger with
  experience, and finally become uncompetitive due to age.  Further
  work might include the introduction of additional reified entities
  that account for such phenomena: we might have a ``mid career
  monster'' who helps competitors at or near their career peak.}

>>>>

\textcolor{red}{One strong assumption made in the
  Kasparov-Karpov-Anand analysis was the constancy of the players'
  strengths over a 36-year period.  One approach might be to consider
  only matches played over a shorter span, but this would result in a
  smaller and less informative dataset; the tension between timeliness
  and informativeness is a delicate and difficult matter.}

  \textcolor{orange}{In a wider context, many sports have a career arc
    whereby competitors start their careers weak, become stronger with
    experience, and finally become uncompetitive due to age.  Further
    work might include the introduction of additional reified entities
    that account for such phenomena: we might have a ``mid career
    monster'' who helps competitors at or near their career peak.}
